\textbf{Minimum Array Partition}

### 2023-09-29 paper/paper.tex
Finalize paper and clean up the repo

    \usepackage{amsmath}

@@

@@

@@

@@

All 4 methods are evaluated on a set of 10 representative testcases. Each algorithm is run 5 times on each prolem insatnce, and we record the best, worst, and average results and execution times.

Let us begin with testcases representing the $random$ type. For this type of problems, we can see that SA finds close to optimal solutions at a fraction of the time:

\begin{verbatim}
Testcase: tests/random_n50_p3.in
  BRUTE FORCE
    Results  : BEST=3769       WORST=3769       AVG=3769.00   
    Time (s) : BEST=2.35       WORST=2.55       AVG=2.45      
  GENETIC ALGORITHM (GA) [pop_size=50, num_iters=250, elitism_size=10]
    Results  : BEST=3769       WORST=3769       AVG=3769.00   
    Time (s) : BEST=0.81       WORST=0.81       AVG=0.81      
  SIMULATED ANNEALING (SA)
    Results  : BEST=3769       WORST=3769       AVG=3769.00   
    Time (s) : BEST=0.13       WORST=0.14       AVG=0.14      
  VARIABLE NEIGHBOURHOOD SEARCH (VNS) [num_iters=2500]
    Results  : BEST=3769       WORST=3769       AVG=3769.00   
    Time (s) : BEST=0.47       WORST=0.48       AVG=0.47

Testcase: tests/random_n50_p6.in
  BRUTE FORCE - skipping due to speed
  GENETIC ALGORITHM (GA) [pop_size=50, num_iters=250, elitism_size=10]
    Results  : BEST=1043       WORST=1046       AVG=1045.40   
    Time (s) : BEST=1.62       WORST=1.68       AVG=1.65      
  SIMULATED ANNEALING (SA)
    Results  : BEST=1043       WORST=1046       AVG=1043.60   
    Time (s) : BEST=0.97       WORST=1.04       AVG=1.00      
  VARIABLE NEIGHBOURHOOD SEARCH (VNS) [num_iters=2500]
    Results  : BEST=1043       WORST=1046       AVG=1043.60   
    Time (s) : BEST=3.18       WORST=3.44       AVG=3.28

Testcase: tests/random_n100_p5.in
  BRUTE FORCE - skipping due to speed
  GENETIC ALGORITHM (GA) [pop_size=100, num_iters=500, elitism_size=20]
    Results  : BEST=11024      WORST=11066      AVG=11033.40  
    Time (s) : BEST=5.52       WORST=5.71       AVG=5.63      
  SIMULATED ANNEALING (SA)
    Results  : BEST=11024      WORST=11066      AVG=11037.20  
    Time (s) : BEST=0.60       WORST=0.66       AVG=0.64      
  VARIABLE NEIGHBOURHOOD SEARCH (VNS) [num_iters=10000]
    Results  : BEST=11024      WORST=11066      AVG=11049.20  
    Time (s) : BEST=9.29       WORST=9.43       AVG=9.39 
\end{verbatim}

@@

\begin{verbatim}
Testcase: tests/linear_n20_p4.in
BRUTE FORCE
        Results  : BEST=602        WORST=602        AVG=602.00    
        Time (s) : BEST=4.71       WORST=5.17       AVG=4.99      
    GENETIC ALGORITHM (GA) [pop_size=20, num_iters=100, elitism_size=4]
        Results  : BEST=602        WORST=602        AVG=602.00    
        Time (s) : BEST=0.17       WORST=0.18       AVG=0.17      
    SIMULATED ANNEALING (SA)
        Results  : BEST=602        WORST=602        AVG=602.00    
        Time (s) : BEST=0.32       WORST=0.34       AVG=0.33      
    VARIABLE NEIGHBOURHOOD SEARCH (VNS) [num_iters=400]
        Results  : BEST=602        WORST=608        AVG=603.20    
        Time (s) : BEST=0.18       WORST=0.20       AVG=0.19    
\end{verbatim}

@@

@@

>>>>

    \usepackage{amsmath}
    \usepackage{hyperref}
    \usepackage[margin=1.1in,footskip=0.25in]{geometry}

@@

	Simulated Annealing is a probabilistic optimization method inspired by annealing in metallurgy. It's effective for combinatorial optimization problems with large search spaces. SA starts with an initial solution and uses a temperature schedule to accept worse solutions with a probability.

	During the SA process, small changes are introduced through random movements of block borders, allowing for exploration of the solution space.

@@

	Variable Neighborhood Search (VNS) is an optimization technique that iteratively explores different neighborhoods of solutions. It begins with an initial solution and systematically varies the neighborhood structure to search for better solutions.

	During the VNS process, changes are introduced through random movements of (multiple) block borders.

@@

For the purpose of evaluating our optimization algorithms, we created 3 sets (random, linear, and squared) of 4 testcases of different dimensions. The dimensions can be categorized as:
\begin{description}
  \addtolength{\itemindent}{0.80cm}
  \itemsep0em 
  \item[Tiny] $n=50$, $p=3$ (BF friendy)
  \item[Small] $n=80$, $p=6$
  \item[Medium] $n=100$, $p=4$
  \item[Large] $n=500$, $p=10$
\end{description}

Each optimization algorithm is run 5 times for every testcase, and we record the best, worst, and average results and execution times.

We only test BF on tiny testcases to compare the results from optimization algorithms. We run BF once per testcase, since there is no randomness involved.

@@

Let us begin with testcases representing the $random$ type:

\begin{verbatim}
  Testcase: tests/random_n50_p3.in
    BRUTE FORCE
      Result   : 3769
      Time (s) : 2.46      
    GENETIC ALGORITHM (GA) [pop_size=75, num_iters=450, elitism_size=15]
      Results  : BEST=3769       WORST=3769       AVG=3769.00   
      Time (s) : BEST=2.11       WORST=2.28       AVG=2.20      
    SIMULATED ANNEALING (SA)
      Results  : BEST=3769       WORST=3769       AVG=3769.00   
      Time (s) : BEST=0.28       WORST=0.29       AVG=0.28      
    VARIABLE NEIGHBOURHOOD SEARCH (VNS) [num_iters=2500]
      Results  : BEST=3769       WORST=3769       AVG=3769.00   
      Time (s) : BEST=0.46       WORST=0.48       AVG=0.47

  Testcase: tests/random_n80_p6.in
    BRUTE FORCE - skipping due to speed
    GENETIC ALGORITHM (GA) [pop_size=100, num_iters=600, elitism_size=20]
      Results  : BEST=3964       WORST=4072       AVG=3985.60   
      Time (s) : BEST=7.86       WORST=8.07       AVG=7.94      
    SIMULATED ANNEALING (SA)
      Results  : BEST=3964       WORST=3964       AVG=3964.00   
      Time (s) : BEST=2.18       WORST=2.27       AVG=2.21      
    VARIABLE NEIGHBOURHOOD SEARCH (VNS) [num_iters=6400]
      Results  : BEST=3964       WORST=3964       AVG=3964.00   
      Time (s) : BEST=8.26       WORST=8.45       AVG=8.35

  Testcase: tests/random_n100_p4.in
    BRUTE FORCE - skipping due to speed
    GENETIC ALGORITHM (GA) [pop_size=100, num_iters=600, elitism_size=20]
      Results  : BEST=17061      WORST=17061      AVG=17061.00  
      Time (s) : BEST=5.06       WORST=5.22       AVG=5.11      
    SIMULATED ANNEALING (SA)
      Results  : BEST=17061      WORST=17061      AVG=17061.00  
      Time (s) : BEST=0.68       WORST=0.68       AVG=0.68      
    VARIABLE NEIGHBOURHOOD SEARCH (VNS) [num_iters=10000]
      Results  : BEST=17061      WORST=17061      AVG=17061.00  
      Time (s) : BEST=4.75       WORST=4.85       AVG=4.80

  Testcase: tests/random_n500_p10.in
    BRUTE FORCE - skipping due to speed
    GENETIC ALGORITHM (GA) [pop_size=100, num_iters=600, elitism_size=20]
      Results  : BEST=326181     WORST=338514     AVG=330243.80 
      Time (s) : BEST=24.44      WORST=25.95      AVG=24.95     
    SIMULATED ANNEALING (SA)
      Results  : BEST=360022     WORST=552450     AVG=424930.40 
      Time (s) : BEST=9.21       WORST=10.39      AVG=9.89      
    VARIABLE NEIGHBOURHOOD SEARCH (VNS) [num_iters=10000]
      Results  : BEST=323192     WORST=417608     AVG=362644.00 
      Time (s) : BEST=24.83      WORST=28.67      AVG=26.29   
\end{verbatim}

For this first type of matrices, we can see that SA tends to find optimal (or close to optimal) solutions at a fraction of the time for tiny, small, and medium tests. However, SA falls behind on the large test case, displaying a big difference between its best (360022) and worst (552450) runs.

Both GA and VNS seem clearly better than SA, however our genetic approach has the better average result across all 5 runs, indicating it is much less reliant on chance to find a good solution.

@@

\begin{verbatim}
  Testcase: tests/linear_n50_p3.in
    BRUTE FORCE
      Result   : 15080
      Time (s) : 3.18      
    GENETIC ALGORITHM (GA) [pop_size=75, num_iters=450, elitism_size=15]
      Results  : BEST=15080      WORST=15318      AVG=15127.60  
      Time (s) : BEST=2.15       WORST=2.24       AVG=2.19      
    SIMULATED ANNEALING (SA)
      Results  : BEST=15080      WORST=15080      AVG=15080.00  
      Time (s) : BEST=0.28       WORST=0.30       AVG=0.29      
    VARIABLE NEIGHBOURHOOD SEARCH (VNS) [num_iters=2500]
      Results  : BEST=15080      WORST=15318      AVG=15270.40  
      Time (s) : BEST=0.40       WORST=0.46       AVG=0.42

  Testcase: tests/linear_n80_p6.in
    BRUTE FORCE - skipping due to speed
    GENETIC ALGORITHM (GA) [pop_size=100, num_iters=600, elitism_size=20]
      Results  : BEST=16055      WORST=16055      AVG=16055.00  
      Time (s) : BEST=7.88       WORST=8.10       AVG=7.98      
    SIMULATED ANNEALING (SA)
      Results  : BEST=15936      WORST=16055      AVG=15983.60  
      Time (s) : BEST=2.23       WORST=2.42       AVG=2.33      
    VARIABLE NEIGHBOURHOOD SEARCH (VNS) [num_iters=6400]
      Results  : BEST=16055      WORST=16055      AVG=16055.00  
      Time (s) : BEST=8.13       WORST=8.32       AVG=8.24

  Testcase: tests/linear_n100_p4.in
    BRUTE FORCE - skipping due to speed
    GENETIC ALGORITHM (GA) [pop_size=100, num_iters=600, elitism_size=20]
      Results  : BEST=67760      WORST=68355      AVG=67879.00  
      Time (s) : BEST=5.16       WORST=5.34       AVG=5.23      
    SIMULATED ANNEALING (SA)
      Results  : BEST=67760      WORST=68355      AVG=67879.00  
      Time (s) : BEST=0.71       WORST=0.71       AVG=0.71      
    VARIABLE NEIGHBOURHOOD SEARCH (VNS) [num_iters=10000]
      Results  : BEST=67760      WORST=72500      AVG=69065.00  
      Time (s) : BEST=4.03       WORST=4.87       AVG=4.38

  Testcase: tests/linear_n500_p10.in
    BRUTE FORCE - skipping due to speed
    GENETIC ALGORITHM (GA) [pop_size=100, num_iters=600, elitism_size=20]
      Results  : BEST=1371552    WORST=1621573    AVG=1454183.80
      Time (s) : BEST=23.79      WORST=26.55      AVG=24.81     
    SIMULATED ANNEALING (SA)
      Results  : BEST=1383025    WORST=2073344    AVG=1565182.40
      Time (s) : BEST=9.15       WORST=11.14      AVG=9.76      
    VARIABLE NEIGHBOURHOOD SEARCH (VNS) [num_iters=10000]
      Results  : BEST=1378620    WORST=2168270    AVG=1566251.40
      Time (s) : BEST=21.42      WORST=29.76      AVG=25.80        
\end{verbatim}

Here the story is very similar to what we observed with the previous test type on smaller test cases. Once again SA is able to find a competitive solution in a fraction of the time. In fact, it even beats GA and VNS on the small test, abeit only by a small margin.

Once again, SA has a big difference between its best and worst runs on the large test, similarly to VNS. GA displays the best performance, and the lowest dependance on chance.

The last group of test cases belong to the $squared$ group:

\begin{verbatim}
  Testcase: tests/squared_n50_p3.in
    BRUTE FORCE
      Result   : 520720
      Time (s) : 4.03      
    GENETIC ALGORITHM (GA) [pop_size=75, num_iters=450, elitism_size=15]
      Results  : BEST=520720     WORST=520720     AVG=520720.00 
      Time (s) : BEST=2.13       WORST=2.22       AVG=2.17      
    SIMULATED ANNEALING (SA)
      Results  : BEST=520720     WORST=520720     AVG=520720.00 
      Time (s) : BEST=0.28       WORST=0.29       AVG=0.28      
    VARIABLE NEIGHBOURHOOD SEARCH (VNS) [num_iters=2500]
      Results  : BEST=520720     WORST=525369     AVG=521649.80 
      Time (s) : BEST=0.47       WORST=0.48       AVG=0.47

  Testcase: tests/squared_n80_p6.in
    BRUTE FORCE - skipping due to speed
    GENETIC ALGORITHM (GA) [pop_size=100, num_iters=600, elitism_size=20]
      Results  : BEST=902840     WORST=968898     AVG=916051.60 
      Time (s) : BEST=8.06       WORST=8.19       AVG=8.12      
    SIMULATED ANNEALING (SA)
      Results  : BEST=901215     WORST=922944     AVG=905560.80 
      Time (s) : BEST=2.25       WORST=2.35       AVG=2.30      
    VARIABLE NEIGHBOURHOOD SEARCH (VNS) [num_iters=6400]
      Results  : BEST=902840     WORST=902840     AVG=902840.00 
      Time (s) : BEST=7.94       WORST=8.32       AVG=8.13

  Testcase: tests/squared_n100_p4.in
    BRUTE FORCE - skipping due to speed
    GENETIC ALGORITHM (GA) [pop_size=100, num_iters=600, elitism_size=20]
      Results  : BEST=4808345    WORST=4808345    AVG=4808345.00
      Time (s) : BEST=5.10       WORST=5.16       AVG=5.14      
    SIMULATED ANNEALING (SA)
      Results  : BEST=4807152    WORST=4807152    AVG=4807152.00
      Time (s) : BEST=0.65       WORST=0.70       AVG=0.67      
    VARIABLE NEIGHBOURHOOD SEARCH (VNS) [num_iters=10000]
      Results  : BEST=4808345    WORST=5033479    AVG=4853371.80
      Time (s) : BEST=4.16       WORST=4.76       AVG=4.61

  Testcase: tests/squared_n500_p10.in
    BRUTE FORCE - skipping due to speed
    GENETIC ALGORITHM (GA) [pop_size=100, num_iters=600, elitism_size=20]
      Results  : BEST=522518392  WORST=619537980  AVG=571390189.00
      Time (s) : BEST=23.39      WORST=26.13      AVG=25.28     
    SIMULATED ANNEALING (SA)
      Results  : BEST=533205270  WORST=1038957839 AVG=680165623.40
      Time (s) : BEST=9.39       WORST=11.87      AVG=10.50     
    VARIABLE NEIGHBOURHOOD SEARCH (VNS) [num_iters=10000]
      Results  : BEST=551863221  WORST=851774417  AVG=660668487.00
      Time (s) : BEST=19.00      WORST=25.25      AVG=22.11 
\end{verbatim}

We find the results for $squared$ test the most interesting.

The results are very similar to the previous group when we look at the smaller test dimensions.

Interestingly, GA had one particularly bad run on the small squared test (968898 compared to its 902840 best). Looking at the average (916051.60), we conclude that only 1 out of 5 runs exhibited such behavior. Despite displaying the smallest reliance on chance before, it is important to note that all optimization algorithms do still depend on random number generation, and as such can be prone to individual bad peformances.

The largest test once again exhibits the results we have observer beforehand, with GA having the smallest result interval, and the lowest average by a significant margin.

@@

In conclusion, the Minimum Array Partition problem has been established as NP-hard, making it challenging to find optimal solutions efficiently. While there exist polynomial algorithms that can provide approximations within a constant factor, their implementation complexity often outweighs the benefits (our approximations factor is better).

As demonstrated in our study, optimization algorithms can be successfully used to solve the Minimum Array Partition problem, with genetic approach displaying the most consistent results, followed closely by VNS. SA remains as a good alternative for when we need to find a decent solution quickly.

These methods strike a balance between computational efficiency and solution quality, making them valuable tools for addressing the Minimum Array Partition problem.

@@

\href{https://doi.org/10.1007/3-540-63165-8_216}{Efficient array partitioning - Sanjeev Khanna, S. Muthukrishnan \& Steven Skiena} \\
\href{https://doi.org/10.1007/BFb0030123}{On the complexity of the generalized block distribution - Michelangelo Grigni \& Fredrik Manne}
